# Data Cleaning
```{r, include = FALSE}
library(tidyverse)
library(dplyr)
```
```{r, echo = FALSE}
# Function of abbreviating character values in the data set 
abbreviate <- function(ds, total = FALSE, race = FALSE){
    if (!race){
        ds$Education[which(ds$Education == "Less.than.high.school.graduate")] <- "Less Than High School"
        ds$Education[which(ds$Education == "High.school.graduate..includes.equivalency.")] <- "High School Graduate"
        ds$Education[which(ds$Education == "Some.college.or.associate.s.degree")] <- "Some Colleges"
        ds$Education[which(ds$Education == "Bachelor.s.degree")] <- "Bachelor Degree"
        ds$Education[which(ds$Education == "Graduate.or.professional.degree")] <- "Advanced Degree"
    }
    
    if (race){
        ds$Education[which(ds$Education == "High.school.graduate.or.higher")] <- "High School or Higher"
        ds$Education[which(ds$Education == "Bachelor.s.degree.or.higher")] <- "Bachelor Degree or Higher"
        ds$Race[which(ds$Race == "White.alone..not.Hispanic.or.Latino" | ds$Race == "White.alone.not.Hispanic.or.Latino")] <- "White Alone"
        ds$Race[which(ds$Race == "Hispanic.or.Latino.Origin")] <- "Hispanic or Latino Origin"
        ds$Race[which(ds$Race == "Black.alone")] <- "Black Alone"
        ds$Race[which(ds$Race == "Asian.alone")] <- "Asian Alone"
        ds$Race<- as.factor(ds$Race)
        ds$Race<-fct_relevel(ds$Race,"White Alone","Hispanic or Latino Origin","Black Alone", "Asian Alone","Other")

    }
    ds$Education <- as.factor(ds$Education)
    if(total){
      ds$Education <- fct_relevel(ds$Education,"Less Than High School","High School Graduate","Some Colleges","Bachelor Degree", "Advanced Degree","Total")
    }
    if(!total&&!race){
      ds$Education<-fct_relevel(ds$Education,"Less Than High School","High School Graduate","Some Colleges","Bachelor Degree", "Advanced Degree")

    }
    if(race){
      ds$Education<-fct_relevel(ds$Education,"Total","High School or Higher",">= Bachelor Degree or Higher")
    }
    return(ds)
}
```

The data sets that we gained from the official census website are in the forms of csv files before the transformation. Because of the unique characteristic of the census data, which are all count values with long descriptions, we primarily focus on transforming and pre-processing the data set into the version that we could apply in visualizing and analysis. It mainly contains the following steps: 

1. Read csv files as dataframes
2. Transform the 'messy' data set into a 'tidy' one
3. Formalize and abbreviate the categorical values 
4. Combine some groups (with relatively few frequencies) into one group

Let's use one of our data sets, 'states' as an example. After we read the csv files into data frame in R and extract the columns that we primarily focus on, the data contains multiple columns indicating education levels, with long and complicated names. 
```{r}
states <- read.csv("./data/states.csv", na.strings = "N/A")
#extract columns for education attainment
states_education <- states[,1:7]
```
```{r, echo=FALSE}
print(colnames(states_education))
```


To have a tidy data set, we transform those columns indicating the education levels into one column: 
```{r}
states_education <- states_education %>% pivot_longer(cols =!State ,names_to = "Education",values_to = "Population")
```
```{r, echo=FALSE}
print(head(states_education, 5))
```


However, as we can see from the above subset of the dataset, some values in the 'Education' columns are long and complicated. It might create some issues for the following data visualization. Therefore, we abbreviate those values into shorter names. Their full names are indicated in the footnote of each graph using those abbreviations for clarification purpose. 
```{r}
states_education <- abbreviate(states_education,total = TRUE)
head(states_education)
```

The above Step 1-3 are repeated for each data set we use. For some data sets containing multiple categories, some of which have relativel low frequencies, we combine those categories as one group. For instance, in the data set counting the number of people with different education attainments in various races, we combine the race groups like 'Native Hawaiian and Other Pacific Islander alone', 'American Indian or Alaska Native alone', and other race groups into one big group called 'Other'. 

```{r, echo = FALSE, include=FALSE}
states <- read.csv("./data/states.csv", na.strings = "N/A")

#extract columns for education attainment
states_education <- states[,1:7]
states_education <- states_education %>% pivot_longer(cols =!State ,names_to = "Education",values_to = "Population")

#extract columns for median earning
states_earning <- states[,c(1,8,9,10,11,12,13)]
states_earning <- states_earning %>% pivot_longer(cols =!State ,names_to = "Education",values_to = "Earning")

#combine data
states <- cbind(states_education,states_earning[,3])
states <- abbreviate(states,total = TRUE)
```


```{r, echo = FALSE, include=FALSE}
birth <- read.csv("./data/births.csv", na.strings = "N/A")

birth <- birth %>% pivot_longer(cols =1:20 ,names_to = c("Give_birth","Marriage","Education"),
    names_sep = "_",
    values_to = "Estimate")
# Make character values more descriptive and succinct as possible
# Need a note about the inclusion of each group specifically: 
# For example, high school = high school (including equivalency), advanced degree = graduate or professional degree
birth <- abbreviate(birth)
```


```{r, echo = FALSE, include=FALSE}
# Function of combining some races into 'Other Races' category
combine_races <- function(race, combine_list){
    race[race$Race %in% combine_list, "Race"] <- "Other"
    combine_race <- race %>% 
        filter(Race == "Other") %>% 
        group_by(Gender, Education) %>% 
        summarise(Estimate = sum(Estimate), .groups = "drop") %>% 
        mutate(Race = "Other")
    race <- race[!race$Race == "Other", ]
    race <- rbind(race, combine_race)
    return(race)
}
```


```{r, echo = FALSE,include=FALSE}
race <- read.csv("./data/race.csv", na.strings = "N/A")
race<- race %>% pivot_longer(cols = !Gender ,names_to = c("Race","Education"),
    names_sep = "_",
    values_to = "Estimate")

other_race_list = list("Native.Hawaiian.and.Other.Pacific.Islander.alone", "American.Indian.or.Alaska.Native.alone", "Some.other.race.alone", "Two.or.more.races")
race <- combine_races(race, other_race_list)

# Make character values more descriptive and succinct as possible
# Need a note that: white means white alone does not count Hispanic or Latino
# Other races include: xxx
race <- abbreviate(race, race = TRUE,)

```


```{r, echo = FALSE,include=FALSE}
education <- read.csv("./data/2015-2019Education.csv", na.strings = "N/A")
education<-education %>% pivot_longer(cols = !Gender ,names_to = c("Year","Education"),
    names_sep = "_",
    values_to = "Estimate")
education$Year <- as.integer(gsub("X", "", education$Year))

# Make character values more descriptive and succinct as possible
education <- abbreviate(education)
```


```{r, echo = FALSE,include=FALSE}
earning <- read.csv("./data/2015-2019earnings.csv", na.strings = "N/A")
earning<-earning %>% pivot_longer(cols = !Gender ,names_to = c("Year","Education"),
    names_sep = "_",
    values_to = "Earning")
earning$Year <- as.integer(gsub("X", "", earning$Year))
earning <- abbreviate(earning,total=TRUE)


```


