# Interactive component

```{r}
new <-education %>%group_by(Year, Education) %>%
  summarise(Estimate = sum(Estimate)) %>%
  mutate(prop = Estimate/sum(Estimate))%>%
  ungroup()

```

```{r}
edu2015 <- new$prop[new$Year==2015]
edu2016 <- new$prop[new$Year==2016]
edu2017 <- new$prop[new$Year==2017]
edu2018 <- new$prop[new$Year==2018]
edu2019 <- new$prop[new$Year==2019]
eduD3 <- data.frame(cbind(edu2015,edu2016,edu2017,edu2018,edu2019))
#write.csv(eduD3,"./data/eduD3.csv")
attach(eduD3)
```





<script src="https://d3js.org/d3.v6.js"></script>  <!-- link to D3 library -->



  <input type="checkbox" id="lessHS" value="lessHS">
  <label for="vehicle1"> Less than High School</label><br>
  <input type="checkbox" id="HS" value="HS">
  <label for="vehicle2"> High School Graduate</label><br>
  <input type="checkbox" id ="associate" value="associate">
  <label for="vehicle3"> Some College or Associate Degree </label><br>
  <input type="checkbox" id="bachelor" value="bachelor">
  <label for="vehicle3"> Bachelor's Degree</label><br>
  <input type="checkbox" id="advanced" value="advanced">
  <label for="vehicle3"> Graduate or Professional Degree</label><br>
  


<div id= "plot"></div>

<script>
  var w = 750;
  var h = 550;
  var margin = {top: 25, right: 250, bottom: 25,left: 30};
  var innerWidth = w - margin.left - margin.right;
  var innerHeight = h - margin.top - margin.bottom;
  var lessHS = [[0.1285585,2015], [0.1254705,2016], [0.1201455,2017], [0.1167068,2018], [0.1139115,2019]];
  var HS = [[0.2756469,2015], [0.2717640,2016], [0.2713289,2017], [0.2686960,2018], [0.2689317,2019]];
  var associate = [[0.2897543,2015], [0.2899876,2016], [0.2886797,2017], [0.2885941,2018], [0.2858890,2019]];
  var bachelor = [[0.1901267,2015], [0.1933507,2016], [0.1969944,2017], [0.1998540,2018], [0.2033382,2019]];
  var advanced = [[0.1159136,2015], [0.1194270,2016], [0.1228514,2017], [0.1261490,2018], [0.1279295,2019]];
  
  var xScale = d3.scaleBand()
      .domain(d3.range(2015,2020))
      .range([0, innerWidth]);
      
  var yScale = d3.scaleLinear()
      .domain([0.1, 0.3])
      .range([innerHeight, 0])

  var xAxis = d3.axisBottom()
      .scale(xScale);

  var yAxis = d3.axisLeft()
      .scale(yScale);
  var svg =d3.select("div#plot")
      .append("svg")
      .attr("width",w)
      .attr("height",h);
      
  var mylinegen = d3.line()
  
  mylinegen
    .x(d => xScale(d[1])+50)
    .y(d => yScale(d[0]));
 
  svg.append("g")
      .attr("class", "xAxis")
      .attr("transform", `translate (${margin.left}, ${h - margin.bottom})`)
      .call(xAxis);

  svg.append("g")
      .attr("class", "yAxis")
      .attr("transform", `translate (${margin.left}, ${margin.top})`)
      .call(yAxis);
      
	d3.select("#lessHS").on("change",update1);
	d3.select("#HS").on("change",update2);
	d3.select("#associate").on("change",update3);
	d3.select("#bachelor").on("change",update4);
	d3.select("#advanced").on("change",update5);
  
  function drawline(data,ID,color,svg_){
  	var mypath = mylinegen(data);
    svg_.append("g")
				.attr("id",ID)
        .attr("transform", `translate (${margin.left}, ${margin.top})`)
        .append("path")
        .attr("d", mypath)
        .attr("fill", "none").attr("stroke", color)
        .attr("stroke-width", "3");
  }
  
  function add_legend(text, color,y,ID){
    svg.append("g")
        .attr("id", ID)
        .append("rect")
        .attr("x","500")
        .attr("y",y)
        .attr("width","20")
        .attr("height","20")
        .attr("fill", color);
    svg.append("text")
        .attr("id",ID)
        .attr("x","525")
        .attr("y", y+20)
        .text(text);
  }
	function update1(){
		  if(d3.select("input#lessHS").property("checked")){
		    drawline(lessHS,"lesshs","#c7f3ff",svg);
		    add_legend("Less Than High School","#c7f3ff",25,"legend1");
		  }else{
		    svg.select("#lesshs").remove();
		    svg.select("#legend1").remove();
		    svg.select("text#legend1").remove();
		  }
	};
	
	function update2(){
	  if(d3.select("input#HS").property("checked")){
		    drawline(HS,"hs","#b3cde3",svg);
		    add_legend("High School Graduate","#b3cde3",50,"legend2");
		}else{
		    svg.select("#hs").remove();
		    svg.select("#legend2").remove();
		    svg.select("text#legend2").remove();
		}
	};
	function update3(){
	  if(d3.select("input#associate").property("checked")){
		    drawline(associate,"associated","#8c96c6",svg);
		    add_legend("Some College Degree","#8c96c6",75,"legend3");

		}else{
        svg.select("#associated").remove();
		    svg.select("#legend3").remove();
		    svg.select("text#legend3").remove();
		}
	};
	function update4(){
	  if(d3.select("input#bachelor").property("checked")){
		    drawline(bachelor,"bachelors","#9b6cb9",svg);
		    add_legend("Bachelor's Degree","#9b6cb9",100,"legend4");

		}else{
        svg.select("#bachelors").remove();
		    svg.select("#legend4").remove();
		    svg.select("text#legend4").remove();
		}
	};
	function update5(){
	  if(d3.select("input#advanced").property("checked")){
		    drawline(advanced,"advance","#810f7c",svg);
		    add_legend("Graduate/Professional Degree","#810f7c",125,"legend5");

		}else{
        svg.select("#advance").remove();
        svg.select("#legend5").remove();
		    svg.select("text#legend5").remove();
		
		}
	};

</script>


